### 2015-06-07 chapters/pitfalls.tex
updated pitfall file format

Further, the rendering of a glyph is also always dependent on the font being
used. For example, the Unicode character \textsc{latin small letter g} appears
as <g> and <{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts,
respectively, because their typefaces are designed differently. Furthermore,
font face may change the visual appearance of a character, for example Times New
Roman two-story <a> changes to a single-story a in italics <\emph{a}>. This
becomes a real problem for some phonetic typesetting (see
Section~\ref{pitfall-ipa-homoglyphs}).

@@

Unicode is a character encoding standard, but these characters of course
actually appear inside some kind of file. The most basic Unicode-based file
format is pure line-based text, i.e.~strings of Unicode-encoded characters
separated by line breaks (note that these line breaks are what for most people
intuitively corresponds to paragraph breaks). Unfortunately, even within this
apparently basic setting there exist a multitude of variants. In general, these
different possibilities are well-understood in the software industry, and
nowadays they normally do not lead to any problems for the end user. However,
there are some situations in which a user is suddenly confronted with cryptic
questions in the user interface involving abbreviations like LF, CR, BE, LE or
BOM.\@ Most prominently this occurs with exporting or importing data in several
software applications from Microsoft. Basically, there are two different issues
involved. First, the encoding of line breaks and, second, the encoding of the
Unicode characters into \textsc{code units} and the related issue of
\textsc{endianness}.

@@

The issue with \textsc{line breaks} originated with the instructions necessary
to direct a printing head to a new line. This involves two movements, known as
\textsc{carriage return} (CR, returning the printing head to the start of the
line on the page) and \textsc{line feed} (LF, moving the printing head to the
next line on the page). Physically, these are two different events, but
conceptually together they form one action. In the history of computing, various
encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only
CR). Currently, all Unix and Unix-derived systems use only LF as code for a
line break, while software from Microsoft still uses a combination of CR and
LF.\@ Today, most software applications recognize both options, and are able to
deal with either encoding of line breaks (until rather recently this was not the
case, and using the wrong line breaks could lead to unexpected errors). Our
impression is that there is a strong tendency in software development to
standardize on the simpler \textsc{only LF} encoding for line breaks, and we
suggest that everybody use this encoding whenever possible.

@@

The issue with \textsc{code units} stems from the question how to separate a
stream of binary ones and zero, i.e.~bits, into chunks representing Unicode
characters. A code unit is the sequence of bits used to encode a single
character in an encoding. Depending on different use cases, the Unicode Standard
offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The
letters UTF stand for either \textsc{Unicode Transformation Format} or
\textsc{Universal Coded Character Set Transformation Format}, but the notion of
\textit{transformation} is a legacy notion that does not have meaning anymore.
Nevertheless, the designation \textsc{UTF} (in capitals) has become an official
standard designation, but should probably best be read as simply Unicode
Format.} The details of this issue is extensively explained in section 2.5 of
the Unicode Core Specification~\cite{Unicode2014}.

@@

Finally, \textsc{UTF-8} is a more complex system that dynamically
encodes each character with the minimally necessary number of bits, choosing
either 8, 16 or 32 bits depending on the character. This represents again a
strong reduction in space (particularly due to the high frequency of data using
ASCII characters, which need only 8 bits) at the expense of even more computation
necessary to process such strings. However, because of the ever growing
computational power of modern machines, the processing overhead is in most
practical situations a non-issue, while saving on space is still useful,
particularly for sending texts over the Internet. As an effect, UTF-8 has become
the dominant encoding on the World Wide Web. We suggest that everybody uses
UTF-8 as their default encoding.

A related problem is a general issue about how to store information in computer
memory, which is known as \textsc{endianness}. The details of this issue go
beyond the scope of this book. It suffices to realize that there is a
difference between \textsc{big-endian} (BE) storage and \textsc{little-endian}
(LE) storage. The Unicode Standard offers a possibility to explicitly indicate
what kind of storage is used by starting a file with a so-called \textsc{Byte Order
Mark} (BOM). However, the Unicode Standard does not require the usage of BOM,
preferring other non-Unicode methods to signal to computers which kind of
endianness is used. This issue only arises with UTF-32 and UTF-16 encodings.
When using the preferred UTF-8, using a BOM is theoretically possible, but
strongly dispreferred according to the Unicode Standard. We suggest that
everyone tries to prevent the inclusion of BOM in your data.

>>>>

Further, the rendering of a glyph is dependent on the font being used. For
example, the Unicode character \textsc{latin small letter g} appears as <g> and
<{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts, respectively,
because their typefaces are designed differently. Furthermore, font face may
change the visual appearance of a character, for example Times New Roman
two-story <{\fontspec{Times New Roman}a}> changes to a single-story a in italics
<\emph{\fontspec{Times New Roman}a}>. This becomes a real problem for some
phonetic typesetting (see Section~\ref{pitfall-ipa-homoglyphs}).

@@

Unicode is a character encoding standard, but these characters of course
actually appear inside some kind of computer file. The most basic Unicode-based file
format is pure line-based text, i.e.~strings of Unicode-encoded characters
separated by line breaks (note that these line breaks are what for most people
intuitively corresponds to paragraph breaks). Unfortunately, even within this
apparently basic setting there exist a multitude of variants. In general, these
different possibilities are well-understood in the software industry, and
nowadays they normally do not lead to any problems for the end user. However,
there are some situations in which a user is suddenly confronted with cryptic
questions in the user interface involving abbreviations like LF, CR, BE, LE or
BOM.\@ Most prominently this occurs with exporting or importing data in several
software applications from Microsoft. Basically, there are two different issues
involved. First, the encoding of line breaks and, second, the encoding of the
Unicode characters into code units and the related issue of endianness.

@@

The issue with \textsc{line breaks} originated with the instructions necessary
to direct a printing head of a physical printer to a new line. This involves two
movements, known as \textsc{carriage return} (CR, returning the printing head to
the start of the line on the page) and \textsc{line feed} (LF, moving the
printing head to the next line on the page). Physically, these are two different
events, but conceptually together they form one action. In the history of
computing, various encodings of line breaks have been used (e.g.~CR+LF, LF+CR,
only LF, or only CR). Currently, all Unix and Unix-derived systems use only LF
as code for a line break, while software from Microsoft still uses a combination
of CR+LF.\@ Today, most software applications recognize both options, and
are able to deal with either encoding of line breaks (until rather recently this
was not the case, and using the wrong line breaks would lead to unexpected
errors). Our impression is that there is a strong tendency in software
development to standardize on the simpler ``only LF'' encoding for line
breaks, and we suggest that everybody use this encoding whenever possible.

@@

The issue with \textsc{code units} stems from the question how to separate a
stream of binary ones and zero, i.e.~bits, into chunks representing Unicode
characters. A code unit is the sequence of bits used to encode a single
character in an encoding. Depending on different use cases, the Unicode Standard
offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The
letters UTF stand for \textsc{Unicode Transformation Format}, but the notion of
``transformation'' is a legacy notion that does not have meaning anymore.
Nevertheless, the designation UTF (in capitals) has become an official
standard designation, but should probably best be read as simply ``Unicode
Format.''} The details of this issue is extensively explained in section 2.5 of
the Unicode Core Specification~\citet{Unicode2014}.

@@

Finally, \textsc{UTF-8} is a more complex system that dynamically encodes each
character with the minimally necessary number of bits, choosing either 8, 16 or
32 bits depending on the character. This represents again a strong reduction in
space (particularly due to the high frequency of data using erstwhile ASCII
characters, which need only 8 bits) at the expense of even more computation
necessary to process such strings. However, because of the ever growing
computational power of modern machines, the processing overhead is in most
practical situations a non-issue, while saving on space is still useful,
particularly for sending texts over the Internet. As an effect, UTF-8 has become
the dominant encoding on the World Wide Web. We suggest that everybody uses
UTF-8 as their default encoding.

A related problem is a general issue about how to store information in computer
memory, which is known as \textsc{endianness}. The details of this issue go
beyond the scope of this book. It suffices to realize that there is a
difference between \textsc{big-endian} (BE) storage and \textsc{little-endian}
(LE) storage. The Unicode Standard offers a possibility to explicitly indicate
what kind of storage is used by starting a file with a so-called \textsc{byte order
mark} (BOM). However, the Unicode Standard does not require the usage of BOM,
preferring other non-Unicode methods to signal to computers which kind of
endianness is used. This issue only arises with UTF-32 and UTF-16 encodings.
When using the preferred UTF-8, using a BOM is theoretically possible, but
strongly dispreferred according to the Unicode Standard. We suggest that
everyone tries to prevent the inclusion of BOM in your data.
